SparseIDS: Learning Packet Sampling with Reinforcement Learning

### 2020-01-21 paper_rl/paper.tex
Add paper

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[conference]{IEEEtran}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Explainability and Adversarial Robustness for RNNs}

\author{\IEEEauthorblockN{Alexander Hartl,
Maximilian Bachl, Joachim Fabini, Tanja Zseby}
\IEEEauthorblockA{Technische Universität Wien\\
firstname.lastname@tuwien.ac.at}}

@@

\thispagestyle{plain}
\pagestyle{plain}

\newacronym{ml}{ML}{Machine Learning}
\newacronym{dl}{DL}{Deep Learning}
\newacronym{aml}{AML}{Adversarial Machine Learning}
\newacronym{ids}{IDS}{Intrusion Detection System}
\newacronym{rnn}{RNN}{Recurrent Neural Network}
\newacronym{fgsm}{FGSM}{Fast Gradient Sign Method}
\newacronym{cw}{CW}{Carlini-Wagner method}
\newacronym{pgd}{PGD}{Projected Gradient Descent}
\newacronym{pdp}{PDP}{Partial Dependence Plot}
\newacronym{ars}{ARS}{Adversarial Robustness Score}
\newacronym{ttl}{TTL}{Time-to-Live}
\newacronym{dos}{DoS}{Denial-of-Service}
\newacronym{iat}{IAT}{Interarrival time}

@@

\glspl{rnn} yield attractive properties for %performing intrusion detection on network data.
constructing \glspl{ids} for network data.
With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} vulnerabilities for profit. Recently developed adversarial \gls{ml} techniques focus on computer vision and their applicability to network traffic is not straightforward: Network packets expose fewer features than an image, are sequential and impose several constraints on their features. %For example, destination ports cannot be modified since that could impair communication.

We show that despite these completely different characteristics, adversarial samples can be  generated reliably for \glspl{rnn}.
To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
for sequential data. Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers. Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}. %as a countermeasure. % AH. the score alone is not yet a countermeasure
%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
%which significantly and successfully reduces the attack surface and improves \gls{ars}.

@@

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content. More recently, with the increasing deployment of encryption, %available features have changed. Since payload is no longer available,
the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.

Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties. %, between two applications potentially being located at different physical locations.
When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. % as well as their size. %AH: don't see why pkt size is not available without flows
%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates.

However, for practical use, high detection accuracy is not enough. With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated. For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common network \glspl{ids}.
In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e. minimally different attack flows which are classified as benign. This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision. Our scenario significantly deviates from computer vision because \begin{enumerate*}
\item the number of features is significantly smaller and
\item only certain features can be manipulated if the flow should remain valid. % (for example the destination port cannot be manipulated)
\end{enumerate*}

Surprisingly, we can confirm that adversarial samples can be found even when considering these tight constraints. % mentioned before.
Thus, we argue that traditional performance metrics like, e.g., accuracy are not sufficient for evaluating security-sensitive \gls{ml} systems.
We propose the \acrfull{ars} as a new performance score, which captures the notion of how easily an adversary can generate adversarial samples.

Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans. This recently stirred up increased interest in explainability methods. Also in this case, common methods are designed to work with images or tabular data, but not with sequences. Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. %These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
Astonishingly, feature importance methods reveal that
features, which are manipulated for successful adversarial flows,
%the that are manipulable features, that are left for modification by adversaries
are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.

%Since common explainability methods are designed to work with images or tabular data but not sequences, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.

Based on the insights gained by the feature importance and explainability methods, we propose two defense methods.
%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
by deploying an adversarial training procedure,
%AH. compared to other research about adversarial training, I dont feel that out approach is so special
%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy
we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.

We make all the source code, data and figures freely available to enable reproducibility and encourage experimentation at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.

\section{An \gls{rnn}-based Classifier}

@@

\begin{table}[b]
\caption{Flow occurrence frequency of attack types.}
\label{tab:occurrence}
\centering
\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
DoS Hulk	&	10.10\%	\\
PortScan, Firewall	&	6.90\%	\\
DDoS LOIT	&	4.08\%	\\
Infiltration	&	3.30\%	\\
DoS GoldenEye	&	0.32\%	\\
DoS SlowHTTPTest	&	0.18\%	\\
DoS Slowloris	&	0.17\%	\\
Brute-force SSH	&	0.11\%	\\
Botnet ARES	&	0.03\%	\\
XSS attack	&	0.03\%	\\
PortScan, no Fw.	&	0.02\%	\\
Brute-force FTP	&	0.01\%	\\
SQL injection	&	$<$0.01\%	\\
Heartbleed	&	$<$0.01\%	\\
\bottomrule
\end{tabular}
}{}
\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
Exploits	&	1.42\%	\\
Fuzzers	&	1.01\%	\\
Reconnaissance	&	0.58\%	\\
Generic	&	0.21\%	\\
DoS	&	0.19\%	\\
Shellcode	&	0.08\%	\\
Analysis	&	0.03\%	\\
Backdoors	&	0.02\%	\\
Worms	&	0.01\%	\\
\bottomrule
\end{tabular}
}{}
\end{table}

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
\autoref{tab:performance_results} shows that our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}. However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows terminate.

\begin{table}
% Evaluated using
% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
% on Dec. 4th
\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
\centering
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
\bottomrule
\end{tabular}
\end{table}

\section{Adversarial Attacks}
\label{sec:adv}
We now investigate whether known \gls{aml} methods can be used for generating adversarial flows for our \gls{rnn}. \cite{hashemi_towards_2019} previously studied the behavior of several Network \glspl{ids} under adversarial attacks but unlike us did not investigate \glspl{rnn}.

A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as TCP flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail. We identify the packet length and the \gls{iat} as features which are most likely to be exploited and thus choose them to be modified by the adversary. But even these features cannot be manipulated due to problem-specific constraints:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Only packets can be manipulated which are transmitted by the attacker, except for botnet and backdoor traffic, which is entirely controlled by an attacker and thus only packets travelling in one direction can be manipulated.
 %This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted.
\item \glspl{iat} must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. An in-depth analysis of cases in which reduction of \glspl{iat} is legitimate is complex, so we generally disallowed \gls{iat} reductions.
\end{itemize}

Several \gls{aml} methods have been proposed in the recent literature, achieving different speed-quality tradeoffs:
\cite{szegedy_intriguing_2014} shows that it is possible to create adversarial samples for an image, which look similar to the original image but are classified wrongly. \cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible. \cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods. %, which we consider in this paper.
% Max: But FGSM is high-speed... So I removed this paragraph altogether
%As finding adversarial flows is particularly difficult in our case, % for the reasons we discussed above,
%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods.
We implemented the following \gls{aml} methods.

\subsection{Carlini-Wagner}
We implemented the \gls{cw} \cite{carlini_towards_2017}, performing gradient descent on the optimization objective
\begin{equation} \label{eq:carliniWagner}
d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta).
\end{equation}
Here, $d(\cdot)$ is a distance metric and $\kappa \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow. Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow optimized by \gls{cw}.

$\delta \in \mathbb R$ is a parameter that determines how far an adversary wants to exceed the decision boundary.
In the original publication $\delta=0$, meaning that the network's decision for adversarial samples is just between attack and benign traffic. In the present context, we need to make sure that the classifier's prediction would actually be benign, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim. Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign after the sigmoid activation function.

We used $L_1$ as distance metric, as we consider $L_1$ distance to represent practically relevant differences of network flows best. We used \gls{pgd} for meeting the real-world constraints discussed above.

\subsection{$L_\infty$-bounded Projected Gradient Descent}
\cite{madry_towards_2018} uses a method for generating adversarial samples which deploys \gls{pgd}  to maximize the network's loss function while constraining the achieved $L_\infty$ distance from the original samples. Hence, an important difference to \gls{cw} is that for this method the network's loss instead of its logit output is considered.
We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close to the original samples.

\subsection{Fast Gradient Sign Method}
Finally, we also tested the \gls{fgsm} % which was initially proposed in
 \cite{goodfellow_explaining_2015}, which is the first method for generating adversarial samples. FGSM can be considered a single pass of \gls{pgd} on the loss function with an equality constraint on the $L_\infty$ distance, i.e. the adversarial sample is found as
\begin{equation}
\tilde X = X + \epsilon \, \text{sgn}( \nabla_X L(X)),
\end{equation}
where $L(X)$ denotes the network's loss function and $\epsilon$ denotes the achieved $L_\infty$ distance.

\begin{figure}[h]
\includegraphics[width=\columnwidth]{./plots/adv_comparison/adv_comparison_17.pdf}
\caption{Attack success ratios for both datasets and per attack type for CIC-IDS-2017. Depicted are flow detection accuracies for adversarial flows and for unmodified attack flows.}
\label{fig:adv_per_family}
\end{figure}

\subsection{Evaluation}
\autoref{fig:adv_per_family} depicts the performance for the investigated algorithms for both datasets and for each attack type in CIC-IDS-2017.
We used \gls{cw} with a tradeoff of 1 and, to provide a fair comparison, for \gls{fgsm} and \gls{pgd} we used an average $L_\infty$ distance as observed for \gls{cw}.

%As expected,
\gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
The figure shows significant differences for detecting different attack families in the first place. Also for generating adversarial flows some attack families are more susceptible than others. However, the results match to a high degree with our expectations. For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.

Interestingly, any increase of the $L_\infty$ bound for \gls{pgd} did not yield significantly improved performance. 
\gls{cw} thus generally achieved superior results to $L_\infty$-bounded \gls{pgd}
and we can confirm the recommendation by \cite{carlini_towards_2017} to perform gradient descent on the logit output instead of the loss for good results.

\autoref{fig:adv_cw} depicts the success ratio and average distance for \gls{cw} for different tradeoffs $\kappa$.
%\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as benign with respect to the final packet along with the resulting distance from the original sample.
Evidently, the larger $\kappa$, the better the attack works, but also distances from original flows become higher. With acceptable distances, we were able to generate adversarial samples for about 80\% of attack samples.

%Note, however, that f
For successful adversarial samples, the second term in \autoref{eq:carliniWagner} becomes constant and, hence, is no longer relevant for optimization. Hence, $\kappa$ is irrelevant for the achieved distance, but governs if for one sample an adversarial sample is found.
However, when using a large $\kappa$ in \autoref{eq:carliniWagner}, to avoid overly large steps that impede convergence, it is necessary to reduce the gradient descent learning rate. Since the distance term is not affected by $\kappa$, it is furthermore necessary to scale the iteration count inversely with the learning rate. \gls{cw} with a large $\kappa$ therefore needs more time than with a small $\kappa$.

% The figure suggests that a tradeoff value between 0.5 and 1 would be ideal to find adversarial samples with a low enough distance from real samples.
% This conclusion might be misleading, however, as due to the objective function \eqref{eq:carliniWagner} the distance cannot become larger anymore for one sample as soon as an adversarial sample has been found. %, when performing optimization as depicted by the objective function .
% Hence, the depicted distance increase presumably is due to samples for which no adversarial sample could be found for lower $\kappa$ values.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
\caption{Performance of \gls{cw} with different $\kappa$ values.}
\label{fig:adv_cw}
\end{figure}

\section{Adversarial Robustness Metric}

As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}. High recall for a particular attack type does not imply that adversarial samples are hard to find  %are  likely to fail
for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.

Such a metric should be easy to compute and easy to interpret.
While in general adversarial robustness for \gls{ml} models is frequently quantified as the average or median of minimum distances of adversarial samples \cite{carlini_evaluating_2019}, we found no generally agreed-upon metric for \glspl{ids} in the literature. If we consider flows, for which no adversarial sample can be found, to have infinite distance, the median has the advantage of ignoring such unsuccessful samples and outliers. On the other hand, the median might depict expected distances badly if they have a very uneven distribution. 
% We propose the \gls{ars} as 
% \begin{equation}
% \text{ARS} = \text{mean}(\{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}),
% \text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \{d \in \mathcal D \mid d \leq Q_{0.5}(\mathcal D)\}} d,
% \end{equation}
% denoting by $Q_{0.5}(\cdot)$ the median, by $N = | \mathcal D | \in \mathbb N$ the total number of samples and by $\mathcal D$ the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$.

We thus propose the \gls{ars} as follows. Let $\mathcal S$ denote the set of samples, $N = | \mathcal S | \in \mathbb N$ the total number of samples and $d_s \in \mathbb R_0^+$ the distance of an adversarial flow from the original flow for a sample $s\in \mathcal S$, assigning unsuccessful samples a distance of $\infty$. % and $Q_{0.5}(\cdot)$ the median function.
We define the \gls{ars} as 
\begin{equation}
\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{\tilde s \in \tilde {\mathcal S}} d_{\tilde s},
\end{equation}
with $\tilde {\mathcal S} \subset \mathcal S$ denoting a set of size $|\tilde {\mathcal S}| =\lceil N/2 \rceil$, so that $d_{\tilde s} \le d_s$ for all $\tilde s \in  \tilde {\mathcal S}, s \in   {\mathcal S} \setminus \tilde {\mathcal S} $.
% We thus propose the \gls{ars} as follows. Let $\mathcal D$ denote the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$, $N = | \mathcal D | \in \mathbb N$ the total number of samples and $Q_{0.5}(\cdot)$ the median function.
% We define the \gls{ars} as 
% \begin{equation}
% \text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \tilde {\mathcal D}} d,
% \end{equation}
% with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of size $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
The \gls{ars} is thus approximately the average of distances not larger than the median distance.

%Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
We consider an adversary successful if he can cause at least 50\% of all samples to be misclassified. In this case, the \gls{ars} is the average of the distances of the adversarial samples to the original samples for the 50\% of all samples which have the smallest distances. If the adversary doesn't manage to manipulate enough samples, the \gls{ars} is $\infty$.

\gls{cw} is able to find adversarial samples with minimum distance, but becomes slow for large $\kappa$. Hence, it can be used for finding the \gls{ars} efficiently as follows: 
%Since \gls{cw} becomes slow for large $\kappa$, the \gls{ars} can be efficiently found as follows:
Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If %, of a total of $N$ samples,
 at least $N/2$  are wrongly classified, and
 the smallest distance of non-adversarial samples is not smaller than the $\lceil\frac{N}{2}\rceil$th smallest distance of adversarial samples,
stop and compute the \gls{ars}. Otherwise increase $\kappa$ and repeat. If after a predefined number of iterations -- e.g. 100 -- still not more than half of the samples are adversarial, also break and set the \gls{ars} to $\infty$.

The larger the \gls{ars} is, the more robust the model is, as then an adversary needs to modify the samples more to make them adversarial. If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples. In this case, the ratio of samples that could be made adversarial can be a useful metric to determine the exact extent of the adversarial threat.

Setting the threshold for the \gls{ars} to 50\% is arbitrary, but
 % reasonable as this way the \gls{ars} approximates the median of the distances of all samples and, additionally, marks the distance where an adversarial attack is more likely to be successful than unsuccessful.
 reasonable as outliers with very large distances are ignored and because if an attacker can make the majority of samples adversarial, we argue that its attack is ``successful''.
%we think that it is reasonable since if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.

\begin{figure}[h]
\includegraphics[width=\columnwidth]{./plots/ars_original.pdf}
\caption{Recall for unmodified flows and \gls{ars} for attacks in CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls below 66\%: The attack does not succeed and the \gls{ars} is $\infty$.}
\label{fig:recall_ars}
\end{figure}

\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker aiming to make them look benign.

\section{Explaining Recurrent Neural Networks}
Having verified the effectiveness of \gls{aml}  for our \glspl{rnn}, we now  investigate how the classifiers come to a decision.
From a naive perspective, one might be tempted to reuse existing explainability methods for \glspl{rnn} by considering a flow the sum of its packet features.
We identify several difficulties which occur when trying to explain decisions made by \glspl{rnn}.

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[conference]{IEEEtran}
%% IEEE CNS addition:
\makeatletter
\def\ps@headings{%
\def\@oddhead{\mbox{}\scriptsize\rightmark \hfil \thepage}%
\def\@evenhead{\scriptsize\thepage \hfil \leftmark\mbox{}}%
\def\@oddfoot{}%
\def\@evenfoot{}}
\makeatother
\pagestyle{empty}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{SparseIDS: Learning Packet Sampling with Reinforcement Learning}

\author{\IEEEauthorblockN{Maximilian Bachl, Fares Meghdouri, Joachim Fabini, Tanja Zseby}
\IEEEauthorblockA{Technische Universität Wien\\
firstname.lastname@tuwien.ac.at}}

@@

%\thispagestyle{plain}
%\pagestyle{plain}

\newacronym{ml}{ML}{Machine Learning}
\newacronym{dl}{DL}{Deep Learning}
\newacronym{aml}{AML}{Adversarial Machine Learning}
\newacronym{ids}{IDS}{Intrusion Detection System}
\newacronym{rnn}{RNN}{Recurrent Neural Network}
\newacronym{fgsm}{FGSM}{Fast Gradient Sign Method}
\newacronym{cw}{CW}{Carlini-Wagner method}
\newacronym{pgd}{PGD}{Projected Gradient Descent}
\newacronym{pdp}{PDP}{Partial Dependence Plot}
\newacronym{ars}{ARS}{Adversarial Robustness Score}
\newacronym{ttl}{TTL}{Time-to-Live}
\newacronym{dos}{DoS}{Denial-of-Service}
\newacronym{iat}{IAT}{Interarrival time}
\newacronym{rl}{RL}{Reinforcement Learning}

\newcommand{\ours}{SparseIDS}

@@

\glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data. They allow analyzing network flows and already determining if a flow is malicious or not before it is over, making it possible to take action immediately. However, considering the large number of packets that have to be inspected, the question of computational as well as energy efficiency arises. We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high. Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy with the same number of sampled packets. A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of packets like current approaches but can possibly skip arbitrarily many packets, saving even more computational resources for long sequences. Finally we build an automatic steering mechanism that can guide \ours{} to achieve a desired level of sparsity.

@@

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content. More recently, with the increasing deployment of encryption, the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.

Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties. When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.

However, for practical use, not only classification accuracy and security are important but also computational efficiency. This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding additional security breaches is higher than the cost of the \gls{ids} itself. The cost of the \gls{ids} can be split in the cost of purchase/installation, the cost of operating it and the cost of maintaining it. The goal we pursue here is to develop sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data. For this purpose, we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} that has a tunable parameter that allows trading off accuracy for sparsity of the data. We show that a significant number of packets can be skipped while accuracy doesn't drop significantly. Compared to other common sampling strategies, \ours{} performs better when being trained on the same number of packets.

As \ours{} uses a parameter to trade off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter. However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.

\section{Setup}

@@

\begin{table}[b]
\caption{Flow occurrence frequency of attack types.}
\label{tab:occurrence}
\centering
%\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
%]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
DoS Hulk	&	10.10\%	\\
PortScan, Firewall	&	6.90\%	\\
DDoS LOIT	&	4.08\%	\\
Infiltration	&	3.30\%	\\
DoS GoldenEye	&	0.32\%	\\
DoS SlowHTTPTest	&	0.18\%	\\
DoS Slowloris	&	0.17\%	\\
Brute-force SSH	&	0.11\%	\\
Botnet ARES	&	0.03\%	\\
XSS attack	&	0.03\%	\\
PortScan, no Fw.	&	0.02\%	\\
Brute-force FTP	&	0.01\%	\\
SQL injection	&	$<$0.01\%	\\
Heartbleed	&	$<$0.01\%	\\
\bottomrule
\end{tabular}
%}{}
%\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
%]{
%\begin{tabular}{l r}
%\toprule
%Attack type & \hspace*{-4mm}Proportion \\ \midrule
%Exploits	&	1.42\%	\\
%Fuzzers	&	1.01\%	\\
%Reconnaissance	&	0.58\%	\\
%Generic	&	0.21\%	\\
%DoS	&	0.19\%	\\
%Shellcode	&	0.08\%	\\
%Analysis	&	0.03\%	\\
%Backdoors	&	0.02\%	\\
%Worms	&	0.01\%	\\
%\bottomrule
%\end{tabular}
%}{}
\end{table}

We implemented a three-layer LSTM-based classifier with 128 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

%For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. 
For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.

Attacks contained in the datasets are shown in \autoref{tab:occurrence}.

%\begin{table}
%% Evaluated using
%% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
%% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
%% on Dec. 4th
%\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
%\centering
%\begin{tabular}{l r r r r} \toprule
%& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
%	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
%Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
%Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
%Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
%F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
%Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
%\bottomrule
%\end{tabular}
%\end{table}
